---
title: "Week 3: Wednesday Solutions"
author: Harald Kugler
---


```{r}
#| message: false
#| warning: false
library(ggplot2)
library(tibble)
library(broom)
library(splines)
library(gridExtra)
```

## Exercise 7.1

This is a continuation of Exercise 2.1. We simulate the data set.

```{r}
set.seed(04022016)
factor_labels <- rep(c("A", "B", "C"), each = 50)
formula_data <- tibble(
    x =  rnorm(150, 0),
    z = factor(sample(factor_labels)),
    mu = (z == "A") + cos(x) + 0.5 * (z == "B") * sin(x),
    y = rnorm(150, mu, 0.5)
)
```

#### 1.

We must determine the number of columns of the model matrix for the formula `y ~ ns(x, df = 4) * z` and interpret them. Recall that in a formula the operator `*` expands as
$$
b * c = b + c + b:c,
$$
so the formula includes the main effects of the basis expansion and of the factor as well as all their interactions. The columns are therefore

- $1$ column for the intercept,
- $4$ columns for the natural cubic spline basis functions of $x,$
- $2$ columns for the dummy variable encoding of `z`, which has $3$ levels,
- $2 \cdot 4 = 8$ columns for the interactions between the two dummy variables and the four basis functions,

for a total of $15$ columns. The interaction columns are what allow the nonlinear effect of $x$ to differ between the three groups: within group $A$ (the base level) the effect of $x$ is given by the four main effect coefficients alone, while for groups $B$ and $C$ the corresponding interaction coefficients are added on top.

```{r}
head(model.matrix(y ~ ns(x, df = 4) * z, formula_data))
ncol(model.matrix(y ~ ns(x, df = 4) * z, formula_data))
```

#### 2.

We fit the model and reproduce the figure. The left panel shows the data, the fitted values as solid lines and the true means as dashed lines. The right panel is a residual plot with a smooth added.

```{r}
#| fig-width: 10
#| fig-height: 5
#| message: false
modelSplines <- lm(y ~ ns(x, df = 4) * z, data = formula_data)

p1 <- ggplot(formula_data, aes(x, y, color = z)) +
  geom_point() +
  geom_line(aes(y = mu), linetype = 2) +
  geom_line(aes(y = fitted(modelSplines))) +
  theme_bw() +
  theme(legend.position = "top")

p2 <- ggplot(augment(modelSplines), aes(.fitted, .resid, color = z)) +
  geom_point() +
  geom_smooth(aes(x = .fitted, y = .resid), color = "blue") +
  xlab("Fitted values") +
  ylab("Residuals") +
  theme_bw() +
  theme(legend.position = "top")

grid.arrange(p1, p2, ncol = 2)
```

The fitted curves follow the true means closely within each of the three groups, and in particular the model captures that the relation between $x$ and $y$ is nonlinear and that its shape differs between the groups. The residual smooth is essentially flat around zero, so there is no indication that the mean value assumption **A1** is violated, and the spread of the residuals looks roughly constant across the range of fitted values, so **A2** is plausible as well.

## Exercise 7.2

Consider the following matrices

```{r}
X <- splines::bs(
    seq(0, 1, 0.01),
    knots = c(0.3, 0.5, 0.7)
)
X_prime <- seq(0, 1, 0.01)
```

of $B$-spline basis functions evaluated on a grid in $[0,1].$ By Exercise 2.3, `X_prime` lies in the column space of `X` if there is a $\mathbf{C}$ with $\mathbf{X}' = \mathbf{X}\mathbf{C},$ and in that case

$$
\mathbf{C} = \left(\mathbf{X}^\top\mathbf{X}\right)^{-1}\mathbf{X}^\top\mathbf{X}'.
$$
We can therefore verify the claim by checking that the following expression is numerically zero.

```{r}
max(abs(X %*% solve(t(X) %*% X) %*% t(X) %*% X_prime - X_prime))
```

Since it holds up to numerical error, we know that $\mathbf{X}' = \mathbf{X}\mathbf{C}$ with $\mathbf{C} \in \mathbb{R}^6.$ We can also view this as a regression problem, where $\mathbf{C}$ plays the role of the parameter vector (usually denoted $\beta$) and where the noise is zero. It is then enough to check that all residuals are zero up to numerical error.

```{r}
fit <- lm.fit(X, X_prime)

max(abs(fit$residuals))
# or, equivalently
max(abs(fit$fitted.values - X_prime))
```

Finally, we check that $\mathbf{C}$ really is the parameter vector estimated by `lm.fit()`.

```{r}
C <- solve(t(X) %*% X) %*% t(X) %*% X_prime
max(abs(C - fit$coefficients))
```

